Binding-Time Analysis for \textsc{miniKanren}

### 2020-03-04 papers/TEASE-LP/example.tex
TEASE-LP edits

It has been shown in~\cite{lozov2019relational} that an online conjunctive partial deduction is capable of improving running time of a program in a given direction, but we believe that offline specialization may provide even better results in the context of relational interpreters.
Offline specializers employ a static analysis before the specialization step.
The most prominent of them is \emph{binding-time analysis}.
It is used to determine which data is available statically, and which~--- dynamically.
Having these annotations, a specialization algorithm can ignore dynamic data and only symbolically execute a program with respect to the static data.
This usually leads to a more precise and powerful specialization.

We believe that a variation of a binding-time analysis can serve as a solution to figuring out the order of variable bindings necessary for functional conversion as well as be a part of an offline specializer.
The goal of binding-time analysis is to identify at which step of a computation each variable gets its value.
This process is called annotation: each variable gets an annotation from the domain of \emph{binding-time}.
Existing binding-time analyses for pure \prolog{}~\cite{craig2004fully} and \mercury{}~\cite{vanhoof2004binding} only distinguish between static and dynamic variables, which does not help to work out the order.
Natural numbers serve as annotations in~\cite{Thiemann1997AUF} and are better suited to our problem.

@@

We assume that all goals are in \emph{canonical normal form}: $\underline{fresh} \ V_1 \dots V_n \bigvee \bigwedge (\underline{call} \ R^k \ V_1 \dots V_k \mid V \equiv T)$. A goal is a disjunction of conjunctions of either relation calls or unifications; all free variables are brought up into scope by the $\underline{fresh}$ operator on the top level. Relation calls and constructors can only have variables as their arguments. Unification should have a single variable on the left-hand side with the exception of unifying two constructors of arity $0$.
We leave unifications of the zero-arity constructors out, since they can either be evaluated to a failure or to a success without extending the substitution which can be determined statically.
Any core \miniKanren{} program can be converted to this form.

>>>>

In compilation and specialization to determine at what time the values of the variables can be computed, \emph{binding-time analysis}, or bta, is used.
Typically, bta divides all computations into two stages: \emph{compile-time}, or \emph{static}, and \emph{execution time}, or \emph{dynamic}.
One approach to bta is to explicitly annotate program variables and computations by equipping them with their binding times.
In logic programming there is no predefined execution direction --- no order in which conjunctions should be considered ---
which complicates the problem of identifying variables binding times.
It has been shown~\cite{lozov2019relational} that online conjunctive partial deduction may improve program running time in a given direction.
We are investigating the applicability of offline specialization methods in the context of functional conversion.
Offline approaches are known~\cite{jones1993partial} to be safe, helpful in the design and implementation of partial evaluation projects, and are more efficient than the online ones, which is crucial in the context of logic programming.
The existing binding-time analyses for pure \prolog{}~\cite{craig2004fully} and \mercury{}~\cite{vanhoof2004binding} only distinguish between static and dynamic variables, which does not help to work out the order.
Natural numbers serve as annotations in~\cite{Thiemann1997AUF} and are better suited to our problem.

% It has been shown in~\cite{lozov2019relational} that an online conjunctive partial deduction is capable of improving running time of a program in a given direction, but we believe that offline specialization may provide even better results in the context of relational interpreters.
% Offline specializers employ a static analysis before the specialization step.
% The most prominent of them is \emph{binding-time analysis}.
% It is used to determine which data is available statically, and which~--- dynamically.
% Having these annotations, a specialization algorithm can ignore dynamic data and only symbolically execute a program with respect to the static data.
% This usually leads to a more precise and powerful specialization.

% We believe that a variation of a binding-time analysis can serve as a solution to figuring out the order of variable bindings necessary for functional conversion as well as be a part of an offline specializer.
% The goal of binding-time analysis is to identify at which step of a computation each variable gets its value.
% This process is called annotation: each variable gets an annotation from the domain of \emph{binding-time}.
% Existing binding-time analyses for pure \prolog{}~\cite{craig2004fully} and \mercury{}~\cite{vanhoof2004binding} only distinguish between static and dynamic variables, which does not help to work out the order.
% Natural numbers serve as annotations in~\cite{Thiemann1997AUF} and are better suited to our problem.

@@

We assume that all goals are in \emph{canonical normal form}: a goal is a disjunction of conjunctions of either relation calls or unifications; all free variables are brought up into scope by the $\underline{fresh}$ operator on the top level. Relation calls and constructors can only have variables as their arguments. Unification should have a single variable on the left-hand side with the exception of unifying two constructors of arity $0$.
We leave unifications of the zero-arity constructors out, since they can either be evaluated to a failure or to a success without extending the substitution which can be determined statically.
Any core \miniKanren{} program can be converted to this form.
$$Goal = \underline{fresh} \ V_1 \dots V_n \bigvee \bigwedge (\underline{call} \ R^k \ V_1 \dots V_k \mid V \equiv T)$$
